---
title: "ProBot: Back and Forth Again"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProBot: Back and Forth Again}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ProBot)
library(torch)
library(magicaxis)
set.seed(42)
torch_manual_seed(42)
eval_global <- FALSE
```

## Overview

This vignette demonstrates the full **ProBot** simulation-based inference (SBI) workflow
on a mixture-of-Gaussians simulator — a classic test-bed for likelihood-free inference
methods.  The simulator has five input parameters and returns two low-dimensional summary
statistics (mean and variance of a mixture sample).  Because the summary statistics are a
many-to-one compression of the original data, the inverse problem is genuinely ill-posed:
many different parameter vectors can produce the same (or very similar) summaries.  This
makes it an ideal setting to showcase the power of Mixture Density Networks (MDNs), which
can represent a full *posterior distribution* over plausible inputs rather than a single
point estimate.

The vignette is structured as two complementary inference tasks:

1. **Forward inference** — given the five simulator parameters, predict the two summary
   statistics together with their uncertainty.
2. **Backward (inverse) inference** — given an observed pair of summary statistics,
   recover a posterior distribution over the five parameters.

After training each MDN we validate the posterior calibration using the two diagnostics
built into ProBot:

- **PIT** (Probability Integral Transform) checks *marginal* calibration dimension by
  dimension.
- **TARP** (Test of Accuracy with Random Points) checks *joint* calibration across all
  output dimensions simultaneously.

### The simulator

The mixture-of-Gaussians simulator draws `n_samp` samples from a two-component mixture
and returns the empirical mean and variance of the pooled draw.  Component 1 has mean
`alpha1` and standard deviation `beta1`; component 2 has mean `sqrt(alpha2)` and standard
deviation `sqrt(beta2)`.  The mixing fraction is `gamma`.

```{r simulator}
sample_func <- function(alpha1, beta1, alpha2, beta2, gamma, n_samp = 1e3) {
  n1 <- round(n_samp * gamma)
  n2 <- n_samp - n1
  samp <- c(
    rnorm(n1, mean = alpha1,       sd = beta1),
    rnorm(n2, mean = sqrt(alpha2), sd = sqrt(beta2))
  )
  c(mean = mean(samp), variance = var(samp))
}
```

### Why is this hard to invert?

The five-dimensional parameter space is compressed to just two summary statistics, so the
inverse mapping is inherently degenerate.  Additionally `alpha2` and `beta2` enter the
simulator via square roots, introducing a non-linear and non-injective relationship with
the outputs.  A standard regression network would return a single parameter vector and
silently throw away the multi-modal structure of the posterior; an MDN can represent the
full distribution.

---

## Simulating training and test data

We draw 100 000 parameter vectors uniformly from the prior and evaluate the simulator at
each one to build the training set, and a further 10 000 for testing.

```{r simulate-data, eval = eval_global}
n_train <- 1e5
n_test  <- 1e4

prior_sample <- function(n) {
  cbind(
    alpha1 = runif(n,  1, 10),
    beta1  = runif(n,  1, 10),
    alpha2 = runif(n,  1, 10),
    beta2  = runif(n,  1, 10),
    gamma  = runif(n,  0,  1)
  )
}

params_train <- prior_sample(n_train)
params_test  <- prior_sample(n_test)

stats_train <- t(apply(params_train, 1, function(p) {
  sample_func(p["alpha1"], p["beta1"], p["alpha2"], p["beta2"], p["gamma"])
}))

stats_test <- t(apply(params_test, 1, function(p) {
  sample_func(p["alpha1"], p["beta1"], p["alpha2"], p["beta2"], p["gamma"])
}))

colnames(stats_train) <- colnames(stats_test) <- c("mean", "variance")
```

It is instructive to look at the joint distribution of all seven quantities (five
parameters plus two statistics) before training:

```{r magtri-all, fig.width = 7, fig.height = 7, eval = eval_global}
magtri(cbind(params_train[1:5000, ], stats_train[1:5000, ]))
```

The triangle plot already hints at the compressed nature of the mapping: the summary
statistics have a smooth, low-dimensional relationship with the parameters, but many
parameter combinations are consistent with similar statistic values.

---

## Example 1: forward inference (parameters → summary statistics)

### Scaling and data loading

Neural networks train best when all inputs and outputs are on comparable scales.
`probotScaleForward()` standardises each column to zero mean and unit variance and stores
the centering and scale parameters as attributes so that the same transform can be applied
to new data or reversed on predictions.

```{r forward-prep, eval = eval_global}
params_scaled <- probotScaleForward(params_train)
stats_scaled  <- probotScaleForward(stats_train)

forward_loader <- probotDataLoader(
  input  = params_scaled,
  output = stats_scaled,
  batch  = 1024
)
```

### Building and training the forward MDN

We use a three-layer encoder (64 → 128 → 64 hidden units) with five mixture components.
Five components give the network enough flexibility to represent skewed or multi-modal
predictive distributions without being unnecessarily large.

```{r forward-train, eval = eval_global}
forward_model <- probotMakeMDN(
  input_dim      = ncol(params_scaled),
  output_dim     = ncol(stats_scaled),
  mdn_components = 5,
  hidden_dims    = c(64, 128, 64)
)()

forward_optimizer <- optim_adam(forward_model$parameters, lr = 0.01)

forward_fit <- probotTrainMDN(
  model          = forward_model,
  dataloader     = forward_loader,
  optimizer      = forward_optimizer,
  epochs         = 100,
  mdn_components = 5,
  verbose        = TRUE,
  early_stop     = FALSE,
  stop_window    = 5,
  stop_delta     = 1e-3
)
```

### Inspecting the training history

```{r forward-history, eval = eval_global}
head(forward_fit$history)
tail(forward_fit$history)
```

A monotonically decreasing loss that flattens toward the final epochs is healthy.
Persistent upward steps after an initial descent would indicate overfitting; an early
plateau would suggest a learning rate that is too low.

### Predicting on the test set

We apply the *training* scaling to the test inputs using the stored attributes — using
different scaling at prediction time is a common source of subtle errors.

```{r forward-predict, eval = eval_global}
params_test_scaled <- probotScaleForward(
  params_test,
  col_means = attr(params_scaled, "col_scale")$col_means,
  col_sds   = attr(params_scaled, "col_scale")$col_sds
)

forward_pred <- probotPredictMDN(
  input          = params_test_scaled,
  model          = forward_fit$model,
  mdn_components = 5
)

forward_summary <- probotMarginalPostMDN(
  mdn_output = forward_pred,
  col_means  = attr(stats_scaled, "col_scale")$col_means,
  col_sds    = attr(stats_scaled, "col_scale")$col_sds,
  col_names  = colnames(stats_scaled)
)

round(forward_summary$post_mean[1:5, ], 4)
```

`probotMarginalPostMDN()` collapses the full mixture to per-dimension marginal summaries
(posterior mean and standard deviation) back-transformed to the original scale.

### Overall accuracy

```{r forward-mae, eval = eval_global}
cat("Forward MAE (mean)    :",
    mean(abs(forward_summary$post_mean[, "mean"]     - stats_test[, "mean"])),     "\n")
cat("Forward MAE (variance):",
    mean(abs(forward_summary$post_mean[, "variance"] - stats_test[, "variance"])), "\n")
```

### Truth vs. predicted scatter plots

For each output dimension we plot the posterior mean against the true value, with ±1
posterior standard deviation as error bars.  A tight, symmetric scatter about the
diagonal confirms good forward prediction.

```{r forward-scatter, eval = eval_global}
for (j in seq_len(ncol(stats_test))) {
  magplot(
    stats_test[, j], forward_summary$post_mean[, j],
    xlab = paste0("True ",  colnames(stats_test)[j]),
    ylab = paste0("Predicted ", colnames(stats_test)[j]),
    pch  = "."
  )
  magerr(
    stats_test[, j], forward_summary$post_mean[, j],
    ylo = forward_summary$post_sd[, j],
    col = hsv(v = 0, alpha = 0.05)
  )
  abline(0, 1, col = "red", lty = 2)
}
```

### Posterior samples for one observation

Rather than relying only on marginal summaries we can draw samples from the joint
predictive distribution for a single test observation.  This is the most faithful
representation of what the MDN has learned.

```{r forward-samples, eval = eval_global}
fwd_post_samples <- probotSamplePostMDN(
  input          = params_test_scaled[1, ],
  model          = forward_fit$model,
  mdn_components = 5,
  n_samples      = 2000,
  col_means      = attr(stats_scaled, "col_scale")$col_means,
  col_sds        = attr(stats_scaled, "col_scale")$col_sds,
  col_names      = colnames(stats_train)
)

head(fwd_post_samples)
```

```{r forward-corner, fig.width = 5, fig.height = 5, eval = eval_global}
magtri(fwd_post_samples, refvals = stats_test[1, ])
```

The blue reference lines mark the true summary statistics for observation 1.
Well-calibrated samples should scatter symmetrically around the truth.

### PIT diagnostic: marginal calibration

```{r forward-pit, eval = eval_global}
PIT_forward <- probotPIT(
  input          = params_test_scaled,
  model          = forward_fit$model,
  mdn_components = 5,
  params         = stats_test,
  col_means      = attr(stats_scaled, "col_scale")$col_means,
  col_sds        = attr(stats_scaled, "col_scale")$col_sds,
  col_names      = colnames(stats_scaled)
)
```

```{r forward-pit-plot, fig.width = 5, fig.height = 5, eval = eval_global}
magtri(PIT_forward)
```

Uniform marginal histograms on the diagonal confirm that each output dimension is
individually well calibrated.  A U-shaped histogram would indicate a posterior that is
too narrow (the truth falls in the tails too often); a Λ-shaped histogram would indicate
a posterior that is too wide.

### TARP diagnostic: joint calibration

```{r forward-tarp, eval = eval_global}
TARP_forward <- probotTARP(
  input          = params_test_scaled,
  model          = forward_fit$model,
  mdn_components = 5,
  params         = stats_test,
  col_means      = attr(stats_scaled, "col_scale")$col_means,
  col_sds        = attr(stats_scaled, "col_scale")$col_sds,
  col_names      = colnames(stats_scaled)
)
```

```{r forward-tarp-plot, fig.width = 6, fig.height = 4, eval = eval_global}
maghist(
  TARP_forward, breaks = 20,
  xlab = "TARP coverage", ylab = "Count",
  main = "Forward model: TARP diagnostic"
)
abline(h = length(TARP_forward) / 20, lty = 2, col = "red")

ks.test(TARP_forward, "punif")
```

A flat histogram and a non-significant Kolmogorov–Smirnov p-value indicate a well
calibrated joint posterior.  Excess mass near 0.5 would suggest the posterior is too
wide; excess mass near 0 and 1 would suggest it is too narrow.

---

## Example 2: inverse (backward) inference (summary statistics → parameters)

Now we flip the problem: given the two observed summary statistics, we want to recover a
posterior distribution over the five simulator parameters.

This is the heart of simulation-based inference (SBI).  The mapping from observations to
parameters is many-to-one in the forward direction, so the inverse is many-to-many — the
same pair (mean, variance) is consistent with a large family of parameter vectors.  An
MDN can represent this multimodal, high-dimensional posterior with a single forward pass,
whereas MCMC or nested sampling would require thousands of simulator evaluations for each
new observation.

### Reusing the scaled data (roles swapped)

We reuse the same scaled matrices from the forward example — we simply swap which one is
the input and which is the output.

```{r inverse-prep, eval = eval_global}
inverse_loader <- probotDataLoader(
  input  = stats_scaled,   # observations as inputs
  output = params_scaled,  # parameters as targets
  batch  = 1024
)
```

### Building and training the inverse MDN

The architecture mirrors the forward MDN, but now the input dimension equals the number
of summary statistics (2) and the output dimension equals the number of parameters (5).
Five mixture components allow the network to represent a genuinely multimodal posterior
over parameters — e.g. situations where two distinct parameter configurations produce
indistinguishable summary statistics.

```{r inverse-train, eval = eval_global}
inverse_model <- probotMakeMDN(
  input_dim      = ncol(stats_scaled),
  output_dim     = ncol(params_scaled),
  mdn_components = 5,
  hidden_dims    = c(64, 128, 64)
)()

inverse_optimizer <- optim_adam(inverse_model$parameters, lr = 0.01)

inverse_fit <- probotTrainMDN(
  model          = inverse_model,
  dataloader     = inverse_loader,
  optimizer      = inverse_optimizer,
  epochs         = 100,
  mdn_components = 5,
  verbose        = TRUE,
  early_stop     = FALSE,
  stop_window    = 5,
  stop_delta     = 1e-3
)
```

The only change relative to the forward MDN is the swap of `input` / `output` in
`probotDataLoader()` and the matching swap of `input_dim` / `output_dim` in
`probotMakeMDN()`.  The training loop, diagnostics, and prediction API are identical.

### Training history

```{r inverse-history, eval = eval_global}
head(inverse_fit$history)
tail(inverse_fit$history)
```

The inverse problem is strictly harder than the forward one (we are recovering five
parameters from only two summaries), so we expect a higher final loss reflecting
irreducible posterior uncertainty.

### Predicting the posterior over parameters

```{r inverse-predict, eval = eval_global}
stats_test_scaled <- probotScaleForward(
  stats_test,
  col_means = attr(stats_scaled, "col_scale")$col_means,
  col_sds   = attr(stats_scaled, "col_scale")$col_sds
)

inverse_pred <- probotPredictMDN(
  input          = stats_test_scaled,
  model          = inverse_fit$model,
  mdn_components = 5
)

inverse_summary <- probotMarginalPostMDN(
  mdn_output = inverse_pred,
  col_means  = attr(params_scaled, "col_scale")$col_means,
  col_sds    = attr(params_scaled, "col_scale")$col_sds,
  col_names  = colnames(params_train)
)

head(round(inverse_summary$post_mean, 3))
head(round(inverse_summary$post_sd,   3))
```

The posterior standard deviations (`post_sd`) quantify how much information the two
summary statistics carry about each parameter.  Parameters that enter the simulator
through the same summary statistic will tend to have larger posterior uncertainty and
stronger posterior correlations.

### Truth vs. predicted parameter scatter plots

```{r inverse-scatter, eval = eval_global}
for (j in seq_len(ncol(params_test))) {
  magplot(
    params_test[, j], inverse_summary$post_mean[, j],
    xlab = paste0("True ",      colnames(params_test)[j]),
    ylab = paste0("Predicted ", colnames(params_test)[j]),
    pch  = "."
  )
  magerr(
    params_test[, j], inverse_summary$post_mean[, j],
    ylo = inverse_summary$post_sd[, j],
    col = hsv(v = 0, alpha = 0.05)
  )
  abline(0, 1, col = "red", lty = 2)
}
```

Because the summary statistics compress five degrees of freedom into two, some parameters
will be poorly constrained by the data — those scatter plots will show wide, diffuse
clouds rather than tight diagonals.  This is expected: the MDN faithfully represents the
posterior *including* its broad uncertainty.

### Posterior samples and corner plot for one observation

The corner plot provides the richest view of the posterior — it shows both marginal
distributions (diagonal) and pairwise correlations (off-diagonal panels).  Multi-modal or
banana-shaped distributions, which marginal summaries would miss entirely, are visible
here.

```{r inverse-samples, eval = eval_global}
inv_post_samples <- probotSamplePostMDN(
  input          = stats_test_scaled[1, ],
  model          = inverse_fit$model,
  mdn_components = 5,
  n_samples      = 2000,
  col_means      = attr(params_scaled, "col_scale")$col_means,
  col_sds        = attr(params_scaled, "col_scale")$col_sds,
  col_names      = colnames(params_train)
)

head(inv_post_samples)
```

```{r inverse-corner, fig.width = 7, fig.height = 7, eval = eval_global}
magtri(inv_post_samples, refvals = params_test[1, ])
```

The blue reference lines mark the true generating parameters for observation 1.  Broad
marginal histograms with little mass near the truth are expected for parameters that are
only weakly constrained by the two summary statistics.

### Numerical summary for one observation

```{r inverse-comparison, eval = eval_global}
comparison <- data.frame(
  parameter = colnames(params_test),
  truth     = as.numeric(params_test[1, ]),
  estimate  = as.numeric(inverse_summary$post_mean[1, ]),
  sd        = as.numeric(inverse_summary$post_sd[1,   ])
)
print(comparison, digits = 3)
```

Parameters whose posterior standard deviation is comparable to or larger than the prior
range (e.g. `beta2` if both `alpha2` and `beta2` influence the variance similarly) are
genuinely unidentifiable from the two chosen summaries — the MDN posterior will
approximately reproduce the prior for those dimensions.

### PIT diagnostic: marginal calibration of the inverse posterior

```{r inverse-pit, eval = eval_global}
PIT_inverse <- probotPIT(
  input          = stats_test_scaled,
  model          = inverse_fit$model,
  mdn_components = 5,
  params         = params_test,
  col_means      = attr(params_scaled, "col_scale")$col_means,
  col_sds        = attr(params_scaled, "col_scale")$col_sds,
  col_names      = colnames(params_train)
)
```

```{r inverse-pit-plot, fig.width = 7, fig.height = 7, eval = eval_global}
magtri(PIT_inverse)
```

In the inverse problem the PIT matrix now has five columns (one per parameter).  Uniform
histograms confirm that the marginal posterior for each individual parameter is
well calibrated even though those marginals may be very broad.

### TARP diagnostic: joint calibration of the inverse posterior

```{r inverse-tarp, eval = eval_global}
TARP_inverse <- probotTARP(
  input          = stats_test_scaled,
  model          = inverse_fit$model,
  mdn_components = 5,
  params         = params_test,
  col_means      = attr(params_scaled, "col_scale")$col_means,
  col_sds        = attr(params_scaled, "col_scale")$col_sds,
  col_names      = colnames(params_train)
)
```

```{r inverse-tarp-plot, fig.width = 6, fig.height = 4, eval = eval_global}
maghist(
  TARP_inverse, breaks = 20,
  xlab = "TARP coverage", ylab = "Count",
  main = "Inverse model: TARP diagnostic"
)
abline(h = length(TARP_inverse) / 20, lty = 2, col = "red")

ks.test(TARP_inverse, "punif")
```

A flat TARP histogram and a non-significant KS p-value provide strong evidence that the
joint posterior over all five parameters is well calibrated — a demanding standard that
simultaneously validates the predicted means, uncertainties, and inter-parameter
correlations.  Because the inverse problem is degenerate (five parameters from two
summaries) the joint posterior will be broad, but it should still be *calibrated* — i.e.
the true parameter vector should behave like a random draw from the predicted posterior.

---

## Comparison: forward vs. inverse uncertainty

One of the key insights from running both directions is the asymmetry of uncertainty:

- **Forward** (parameters → statistics): the mapping is a many-to-one compression, but
  for any fixed parameter vector the simulator output is nearly deterministic (dominated
  by Monte Carlo noise from a sample of 1000).  Posterior standard deviations are small.

- **Backward** (statistics → parameters): the mapping is genuinely ill-posed.  The
  posterior standard deviations for poorly identified parameters will be large — sometimes
  comparable to the full prior range.  The MDN represents this honestly rather than
  returning an overconfident point estimate.

This contrast illustrates why SBI and MDNs are so valuable: they provide calibrated
uncertainty estimates that respect the fundamental limits of what can be inferred from
low-dimensional summaries.

---

## Summary

| Function | Role |
|---|---|
| `probotScaleForward()` | Standardise columns to zero mean / unit variance, store scale attributes |
| `probotDataLoader()` | Assemble batched, shuffled training data for torch |
| `probotMakeMDN()` | Define the MDN architecture (hidden layers, mixture components) |
| `probotTrainMDN()` | Optimise the network by minimising negative log-likelihood |
| `probotPredictMDN()` | Run the trained MDN on new inputs to obtain mixture parameters |
| `probotMarginalPostMDN()` | Compute per-dimension posterior means and standard deviations |
| `probotSamplePostMDN()` | Draw joint samples from the predictive mixture distribution |
| `probotPIT()` | Check **marginal** calibration via the Probability Integral Transform |
| `probotTARP()` | Check **joint** calibration via the Test of Accuracy with Random Points |

The workflow is identical for the forward and inverse problems — only the assignment of
"input" and "output" changes.  PIT and TARP together provide a rigorous, two-level
calibration check: PIT catches dimension-wise bias or dispersion errors, while TARP
catches miscalibration in the joint distribution that is invisible to marginal diagnostics
alone.
