---
title: "ProBot: Coupling or AutoReg? Choosing a Flow Architecture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProBot: Coupling or AutoReg? Choosing a Flow Architecture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ProBot)
library(torch)
library(magicaxis)
set.seed(2517)
torch_manual_seed(2517)
# Training cells are slow; set TRUE to run them interactively.
eval_train <- FALSE
```

## Overview

The *Going with the Flow* vignette introduces normalising flows in general and walks
through the coupling (RealNVP) architecture end to end. This vignette answers a
narrower question that comes up every time you call `probotMakeFlow()`:

**should `style` be `"couple"` or `"autoreg"`?**

ProBot ships two conditional flow architectures behind that one facade. They share
the same training loop (`probotTrainFlow()`), the same loss (`probotLossNF()`), the
same sampler (`probotSamplePostNF()`), and the same save/load path (`probotSave()` /
`probotLoad()`). They differ in *how a single layer decides what each parameter
dimension is allowed to see*, and that difference has real consequences for
expressiveness, sampling cost, and numerical behaviour.

Both are trained the same way: maximise

$$
\log p(\theta \mid x) = \log p_{\text{base}}\!\big(f(\theta, x)\big) + \log\left|\det J_f(\theta, x)\right|
$$

Everything below concerns *how* $f$ is built.

---

## The two architectures

### Coupling: `style = "couple"`

`.probotCouplingLayer` splits the parameter vector into two halves and transforms one
half conditioned on the other:

1. Split $\theta = (\theta_1, \theta_2)$ at `d1 = floor(dim_theta / 2)`.
2. A small MLP takes $(\theta_1, x)$ and predicts shift $t$ and log-scale $s$ for $\theta_2$.
3. Apply $z_2 = (\theta_2 - t)\exp(s)$; $\theta_1$ passes through untouched.
4. Swap the halves, so the next layer transforms the complementary subset.

The Jacobian is triangular, so $\log|\det J|$ is just `log_scale summed over the
transformed dimensions`. `probotFlowCouple` stacks `n_layers` of these.

### AutoReg: `style = "autoreg"`

`.probotFlowAutoRegBlock` implements a masked autoregressive flow (MAF). Instead of a
fixed half-split, a degree-based mask enforces a *sequential* dependency structure:

$$
\text{shift}_j,\ \log\text{scale}_j \ \text{depend on}\ \theta_1, \dots, \theta_{j-1}\ \text{and}\ x
$$

The mask is built once at construction time and registered as a buffer, applied to the
weights at every forward pass. The first hidden layer uses a strict `>` comparison so a
hidden node at degree $k$ cannot see $\theta_k$ itself; later layers use `>=` to let the
chain propagate. Conditioning context `x` is given degree 0, so it is visible everywhere.
`probotFlowAutoReg` stacks `n_blocks` such blocks with a `torch_flip` permutation layer
between them, reversing the parameter order so successive blocks condition in opposite
directions.

The distinction is easiest to see by comparing what each layer conditions on for a
four-dimensional parameter vector:

```
coupling (split at d1 = 2)            autoreg (prefix conditioning)
  theta_1, theta_2 pass through         x             -> shift/scale for theta_1
  (theta_1, x) -> transform theta_2    theta_1, x    -> shift/scale for theta_2
  next layer swaps the halves         theta_1:2, x   -> shift/scale for theta_3
                                        theta_1:3, x   -> shift/scale for theta_4
```

Coupling asks "given half the vector, how should I distort the other half?" AutoReg
asks "given everything before this dimension, how should I transform this dimension?"

---

## What they have in common

Both architectures are:

- **Exact-density** — invertible, so $\log p(\theta \mid x)$ is tractable via change of variables.
- **Triangular-Jacobian** — $\log|\det J|$ is a cheap sum of per-dimension log-scales.
- **Soft-clamped** — `soft_clamp` bounds $|s| \le$ `soft_clamp` through $s \leftarrow c\tanh(s/c)$, preventing the log-determinant from running away during training.
- **Trained identically** — same `probotTrainFlow()`, including the gradient clipping it applies (`nn_utils_clip_grad_value_`, clip value 1.0) to both styles.

Because they are interchangeable at the API level, you can train both and compare
without touching any other part of your workflow.

---

## Where they differ

### Parameter count

AutoReg's conditioner must produce shift/log-scale for *every* dimension, for every
degree, from a fully masked dense layer. Coupling's conditioner only produces outputs
for half the dimensions, from a narrower input. The gap widens with `dim_theta`.

```{r param-counts}
n_par <- function(m) sum(vapply(m$parameters, function(p) prod(dim(p)), numeric(1)))

pc <- do.call(rbind, lapply(c(2, 4, 8, 16), function(d) {
  cp <- probotMakeFlow(dim_theta = d, dim_x = 9, n_layers = 6,
                       hidden_dim = 64, device = "cpu")()
  ar <- probotMakeFlow(dim_theta = d, dim_x = 9, n_layers = 6,
                       hidden_dim = 64, device = "cpu", style = "autoreg")()
  data.frame(dim_theta = d, couple = n_par(cp), autoreg = n_par(ar),
             ratio = round(n_par(ar) / n_par(cp), 2))
}))
pc
```

At these settings the architectures are within ~25% of each other on parameter count
even at 16 dimensions — parameter count is *not* the interesting difference.

### Expressiveness per layer

This is the substantive trade-off.

| | Coupling | AutoReg |
|---|---|---|
| What conditions the transform of $\theta_j$ | The *other half* of $\theta$, fixed by `d1` | *All* $\theta_{1..j-1}$, plus $x$ |
| Dependency depth for one dimension | 1 layer per half, alternating | Full prefix, every block |
| Partition | Fixed 50/50 split | Ordered prefix |
| Caveat | Components of the transformed half are updated independently of one another within a layer | Ordering matters; `torch_flip` between blocks mitigates it |

A coupling layer at a single split cannot model within-half dependencies in one pass —
it needs the swap and a second layer. The autoregressive block gets the full prefix in
one pass. So for a fixed layer budget, AutoReg expresses richer dependency structure.
The cost appears immediately at sampling time.

### Sampling cost

The forward pass (training) is fully parallel for both. The **inverse** pass (sampling)
is not: AutoReg must evaluate its masked MLP once per parameter dimension per block,
because $\theta_j$ cannot be computed until $\theta_{1..j-1}$ exist. Coupling inverts in
a single vectorised sweep regardless of dimensionality.

```{r inverse-cost}
inv_cost <- function(d, n = 2000, n_layers = 4, reps = 3) {
  cp <- probotMakeFlow(d, 9, n_layers, 64, device = "cpu")()
  ar <- probotMakeFlow(d, 9, n_layers, 64, device = "cpu", style = "autoreg")()
  z <- torch_randn(c(n, d)); cc <- torch_randn(c(n, 9))
  tc <- min(vapply(seq_len(reps),
                   function(i) system.time(with_no_grad(cp$inverse(z, cc)))[["elapsed"]], numeric(1)))
  ta <- min(vapply(seq_len(reps),
                   function(i) system.time(with_no_grad(ar$inverse(z, cc)))[["elapsed"]], numeric(1)))
  data.frame(dim_theta = d, couple_sec = round(tc, 4),
             autoreg_sec = round(ta, 4), ratio = round(ta / tc, 1))
}

ic <- do.call(rbind, lapply(c(2, 4, 8, 16, 32), inv_cost))
ic
```

Coupling's inverse time is flat in `dim_theta` (overhead-bound at these sizes);
AutoReg's grows roughly linearly, because the work is `n_blocks * dim_theta` masked
network evaluations. The absolute ratios are CPU/overhead-dependent and will vary by
machine and batch size — the *shape of the scaling* is the durable point.

Note that `probotSamplePostNF()` batches the inverse pass, so the penalty is per
*dimension*, not per observation: sampling 10,000 posterior draws for many observations
at once amortises well, whereas sampling one observation at a time in an R loop
compounds it.

### Invertibility in practice

Both architectures round-trip to float32 precision, and for both the error grows with
depth — it is not a differentiator, but it is worth measuring when you push either
architecture to many blocks.

```{r round-trip}
chk <- function(mdl, label) {
  dev <- mdl$parameters[[1]]$device
  tt <- torch_randn(c(500, 4), device = dev)
  cc <- torch_randn(c(500, 9), device = dev)
  with_no_grad({
    f <- mdl$inverse(mdl(tt, cc)$z, cc)
  })
  data.frame(model = label, max_roundtrip_err = (f - tt)$abs()$max()$item())
}

rt <- do.call(rbind, lapply(c(2, 4, 8, 12), function(nl) {
  rbind(
    chk(probotMakeFlow(4, 9, n_layers = nl, hidden_dim = 64, device = "cpu")(), "couple"),
    chk(probotMakeFlow(4, 9, n_layers = nl, hidden_dim = 64, device = "cpu",
                       style = "autoreg")(), "autoreg")
  ) |> transform(n_layers = nl)
}))
rt
```

Errors sit at the 1e-7 to 1e-6 level throughout — the expected noise floor for
float32 arithmetic, rising gently as more transformation steps accumulate. The floor is
backend-dependent: the same coupling round-trip on MPS measured roughly two orders of
magnitude higher than on CPU. Use a round-trip check as a regression tripwire against
your *own* baseline for a given backend rather than against a fixed universal number.

### Default hyperparameters differ

A genuine gotcha. `probotMakeFlow()` defaults to `hidden_dim = 32` and forwards it to
whichever style you choose, but the underlying modules have different defaults:

```{r defaults}
formals(probotFlowCouple)[c("n_layers", "hidden_dim")]
formals(probotFlowAutoReg)[c("n_blocks", "n_layers_per_block", "hidden_dim")]
```

When `probotMakeFlow(style = "autoreg")` is called, `n_layers` is mapped onto
`n_blocks` and `n_layers_per_block` is fixed at 2. Calling `probotFlowAutoReg()`
directly instead gives you `n_blocks = 5` and `hidden_dim = 500` — a substantially
larger model than the facade's defaults.

```{r default-sizes}
n_par <- function(m) sum(vapply(m$parameters, function(p) prod(dim(p)), numeric(1)))

data.frame(
  route = c("probotMakeFlow(style = \"autoreg\")", "probotFlowAutoReg() direct"),
  parameters = c(
    n_par(probotMakeFlow(4, 9, device = "cpu", style = "autoreg")()),
    n_par(probotFlowAutoReg(dim_theta = 4, dim_x = 9, device = "cpu"))
  )
)
```

That is a ~185x difference at the same nominal call shape. If you are comparing the
two styles, set hyperparameters explicitly on both rather than trusting defaults —
otherwise you are benchmarking model size, not architecture.

---

## A worked comparison

The toy problem from *Going with the Flow*: two parameters `(u, v)` generate a
nine-point polynomial curve, and we invert to recover the posterior over `(u, v)`.

```{r toy-data, eval = TRUE}
make_polynomial_output <- function(theta) {
  a0 <- 0.3 + 0.8 * theta[, 1] - 0.4 * theta[, 2]
  a1 <- -1.2 + 0.5 * theta[, 1]^2 + 0.7 * theta[, 2]
  a2 <- 0.6 * sin(pi * theta[, 1]) - 0.3 * theta[, 2]^2
  x_grid <- seq(-1, 1, length.out = 9)
  y <- vapply(x_grid, function(x) a0 + a1 * x + a2 * x^2, numeric(nrow(theta)))
  colnames(y) <- paste0("x", seq_along(x_grid))
  y
}

n_train <- 20000
theta_train <- cbind(u = runif(n_train, -1, 1), v = runif(n_train, -1, 1))
x_train <- make_polynomial_output(theta_train)

#Add a tiny bit of noise to prevent collapse:

x_train[] = rnorm(n = length(x_train),
                mean = x_train,
                sd = diff(range(x_train))*1e-3)

x_sc <- probotScaleForward(x_train)
t_sc <- probotScaleForward(theta_train)
```

Train both styles with an explicit, identical budget — deliberately not the defaults,
for the reason noted above.

```{r train-both, eval = eval_train}
train_style <- function(style, epochs = 200) {
  dl <- probotDataLoader(input = x_sc, output = t_sc, batch = 1024)
  mdl <- probotMakeFlow(dim_theta = 2, dim_x = 9, n_layers = 6,
                        hidden_dim = 64, style = style)()
  t0 <- Sys.time()
  res <- probotTrainFlow(model = mdl, dataloader = dl,
                         optimizer = optim_adam(mdl$parameters, lr = 1e-3),
                         epochs = epochs, verbose = TRUE, early_stop = FALSE)
  list(model = res$model, history = res$history$loss,
       secs = as.numeric(difftime(Sys.time(), t0, units = "secs")))
}

torch_manual_seed(2517)
fit_cp <- train_style("couple")
torch_manual_seed(2517)
fit_ar <- train_style("autoreg")

fit_cp$secs; tail(fit_cp$history, 1)
fit_ar$secs; tail(fit_ar$history, 1)
```

Compare training curves and posterior width on a held-out observation.

```{r compare-plots, fig.width = 8, fig.height = 4, eval = eval_train}
plot(seq_along(fit_cp$history), fit_cp$history, type = "l", col = "steelblue",
     lwd = 2, ylim = range(c(fit_cp$history, fit_ar$history)),
     xlab = "Epoch", ylab = "Mean NLL", main = "Training loss")
lines(seq_along(fit_ar$history), fit_ar$history, col = "darkorange", lwd = 2)
legend("bottomleft", legend = c("coupling", "autoreg"),
       col = c("steelblue", "darkorange"), lwd = 2, bty = "n")
```

```{r posterior-width, fig.width = 8, fig.height = 4, eval = eval_train}
obs <- x_sc[1, , drop = FALSE]
s_cp <- probotSamplePostNF(input = obs, model = fit_cp$model, n_samples = 5000, dim_theta = 2,
  col_means = attributes(t_sc)$col_scale$col_means,
  col_sds = attributes(t_sc)$col_scale$col_sds)
s_ar <- probotSamplePostNF(input = obs, model = fit_ar$model, n_samples = 5000, dim_theta = 2,
  col_means = attributes(t_sc)$col_scale$col_means,
  col_sds = attributes(t_sc)$col_scale$col_sds)

par(mfrow = c(1, 2))
for (j in 1:2) {
  nm <- colnames(theta_train)[j]
  magplot(density(s_cp[, j]), col = 'steelblue', xlab = nm, main = paste("posterior", nm), ylim=c(0,60))
  lines(density(s_ar[, j]), col = 'darkorange', add = TRUE)
  abline(v = theta_train[1, j], col = "red", lwd = 2, lty = 2)
}
legend("topright", legend = c("coupling", "autoreg", "truth"),
       lty = c(1, 1, 2), col = c('steelblue', 'darkorange', "red"), bty = "n")
```

Both have found a reasonable posterior centred around the 'truth' but the autoreg is arguably better.

**A caution on reading these plots.** Do not treat a narrower posterior as better.
Narrower means more confident; whether that confidence is *earned* is a calibration
question, answered by PIT and TARP (see *Going with the Flow* and the CRPS diagnostic
vignette), not by width. In short-budget runs of the comparison above we observed
AutoReg reaching a lower training NLL while producing visibly tighter posteriors — but
at short budgets both styles under-covered held-out truth substantially, so neither
result should be over-interpreted. Judge architectures on calibration diagnostics over
a held-out sample, at a budget where the loss has genuinely plateaued, and compare
wall-clock only on the inverse pass you actually run.

---

## Save, load, and the `flow_style` field

Checkpoints record which architecture they came from, and `probotLoad()` uses it to
rebuild the right skeleton automatically.

```{r save-load}
f <- tempfile(fileext = ".pt")
mdl <- probotMakeFlow(dim_theta = 4, dim_x = 9, n_layers = 3, hidden_dim = 32,
                      device = "cpu", style = "autoreg")()

probotSave(model = mdl, filename = f, model_type = "flow",
           dim_theta = 4, dim_x = 9, n_layers = 3, hidden_dim = 32)

back <- probotLoad(f, device = "cpu")
cat("reloaded class:", paste(class(back$model), collapse = ", "), "\n")
unlink(f)
```

If you are opening a checkpoint written before `flow_style` was recorded, `probotLoad()`
falls back to `"couple"` — which will silently build the wrong skeleton if the weights
are actually autoregressive. Pass `flow_style = "autoreg"` explicitly for those legacy
files, and confirm the reloaded class matches what you trained.

---

## When to use which

**Start with `"couple"`.** It is the package default for defensible reasons: one
vectorised inverse pass regardless of `dim_theta`, cheaper forward and inverse, fewer
parameters, and a flat, predictable sampling runtime. For low-dimensional posteriors
with modest correlation — and for any workflow where you sample in large batches or
interactively — it is usually sufficient.

**Reach for `"autoreg"` when:**

- Your parameters have **strong, asymmetric dependencies** where $\theta_j$ genuinely depends on many of $\theta_{1..j-1}$, and coupling's fixed half-split struggles to express that in the layers you can afford.
- You have **high `dim_theta`** and want richer per-dimension conditioning without stacking many coupling layers.
- The posterior shape is systematically wrong with coupling — banana structures that shear in one direction only, or marginal widths that refuse to tighten while calibration stays healthy.
- You can pay the sampling cost: inverse is ~`n_blocks * dim_theta` masked evaluations, so batch your `probotSamplePostNF()` calls rather than looping over observations in R.

**Stay on `"couple"` when:**

- `dim_theta` is small (roughly single-digit) — the coupling split is not a bottleneck at d = 2 or 3, and AutoReg buys you little.
- You are sampling many observations interactively, or the posterior draws are consumed in a loop you cannot batch.
- You want the shortest path to a stable, reproducible result with fewer moving parts.

The heuristic: **coupling spends expressiveness at training time to buy speed at
sampling time; AutoReg buys expressiveness at the price of a sequential inverse.** If
you are unsure, train both on the same explicit hyperparameters, compare PIT/TARP
calibration on held-out data, and let the diagnostics decide.

---

## Summary

| | `style = "couple"` | `style = "autoreg"` |
|---|---|---|
| Module | `probotFlowCouple()` | `probotFlowAutoReg()` |
| Per-layer conditioning | Other half of `theta` + `x` | Full prefix `theta_{1..j-1}` + `x` |
| Mixing mechanism | Half-swap | `torch_flip` permutation between blocks |
| Forward pass | Parallel | Parallel |
| Inverse pass | Single sweep, flat in `dim_theta` | `n_blocks * dim_theta` masked passes |
| Facade defaults | `n_layers = 4`, `hidden_dim = 32` | `n_blocks` from `n_layers`, `n_layers_per_block = 2` |
| Round-trip error | ~1e-7 to 1e-6 (float32 floor) | ~1e-7 to 1e-6; both grow with depth |
| Best for | Fast sampling, low-d, stable runs | Strong sequential dependence, higher d |
| Watch out for | Half-split misses within-half dependencies | Slow unbatched sampling; ordering effects |
