---
title: "ProBot vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProBot vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ProBot)
library(torch)
set.seed(1)
torch_manual_seed(1)
```

## Overview

ProBot provides a compact interface for training mixture density networks (MDNs) with `torch`. This vignette demonstrates the main workflow on a toy problem where a generative function maps input parameters to the coefficients of a polynomial response, and then shows the inverse problem where we recover plausible inputs from an observed output.

## Example 1: infer the polynomial output from inputs

We define a generative function that turns two input parameters into three polynomial coefficients. Those coefficients are then evaluated on a fixed grid.

```{r data}
make_polynomial_output <- function(theta) {
  a0 <- 0.3 + 0.8 * theta[, 1] - 0.4 * theta[, 2]
  a1 <- -1.2 + 0.5 * theta[, 1]^2 + 0.7 * theta[, 2]
  a2 <- 0.6 * sin(pi * theta[, 1]) - 0.3 * theta[, 2]^2

  x_grid <- seq(-1, 1, length.out = 9)
  y <- vapply(
    x_grid,
    function(x) a0 + a1 * x + a2 * x^2,
    numeric(nrow(theta))
  )

  list(output = t(y), x_grid = x_grid)
}

n_train <- 1500
n_test <- 200

theta_train <- cbind(
  u = runif(n_train, -1, 1),
  v = runif(n_train, -1, 1)
)

theta_test <- cbind(
  u = runif(n_test, -1, 1),
  v = runif(n_test, -1, 1)
)

poly_train <- make_polynomial_output(theta_train)
poly_test <- make_polynomial_output(theta_test)
```

For supervised learning, we standardise the inputs and outputs before creating a data loader.

```{r forward-prep}
input_scaled <- probotScaleForward(theta_train)
output_scaled <- probotScaleForward(poly_train$output)

dataloader <- probotDataLoader(
  input = input_scaled,
  output = output_scaled,
  batch = 256
)
```

Next we build an MDN that predicts the polynomial values directly from the inputs.

```{r forward-train}
forward_model <- probotMakeMDN(
  input_dim = ncol(input_scaled),
  output_dim = ncol(output_scaled),
  mdn_components = 3,
  hidden_dims = c(64, 64)
)()

forward_optimizer <- optim_adam(forward_model$parameters, lr = 0.01)

forward_fit <- probotTrainMDN(
  model = forward_model,
  dataloader = dataloader,
  optimizer = forward_optimizer,
  epochs = 20,
  n_components = 3,
  lambda = 0.05,
  verbose = FALSE,
  early_stop = TRUE,
  stop_window = 5,
  stop_delta = 1e-3
)
```

The training history records the main optimisation diagnostics.

```{r forward-history}
head(forward_fit$history)
```

We can now predict polynomial outputs for new inputs and convert posterior summaries back to the original scale.

```{r forward-predict}
theta_test_scaled <- probotScaleForward(theta_test)
attr(theta_test_scaled, "col_scale") <- attr(input_scaled, "col_scale")
theta_test_scaled <- probotScaleForward(theta_test)
theta_test_scaled <- collapse::TRA(theta_test, attr(input_scaled, "col_scale")$col_means, "-")
theta_test_scaled <- collapse::TRA(theta_test_scaled, attr(input_scaled, "col_scale")$col_sds, "/")

forward_pred <- probotPredictMDN(
  input = theta_test_scaled,
  model = forward_fit$model,
  n_components = 3
)

forward_summary <- probotMarginalPostMDN(
  mdn_output = forward_pred,
  col_means = attr(output_scaled, "col_scale")$col_means,
  col_sds = attr(output_scaled, "col_scale")$col_sds,
  col_names = paste0("x", seq_len(ncol(output_scaled)))
)

round(forward_summary$post_mean[1:3, 1:4], 3)
```

A quick mean absolute error on the test set shows that the model can recover the polynomial output well.

```{r forward-metrics}
mean(abs(forward_summary$post_mean - poly_test$output))
```

## Example 2: invert the mapping with MDN uncertainty

For the inverse problem, the polynomial values are the inputs and the latent parameters are the targets. This is where the MDN is especially useful because multiple parameter settings can explain similar outputs.

```{r inverse-prep}
obs_scaled <- probotScaleForward(poly_train$output)
param_scaled <- probotScaleForward(theta_train)

inverse_loader <- probotDataLoader(
  input = obs_scaled,
  output = param_scaled,
  batch = 256
)

inverse_model <- probotMakeMDN(
  input_dim = ncol(obs_scaled),
  output_dim = ncol(param_scaled),
  mdn_components = 4,
  hidden_dims = c(64, 64)
)()

inverse_optimizer <- optim_adam(inverse_model$parameters, lr = 0.01)

inverse_fit <- probotTrainMDN(
  model = inverse_model,
  dataloader = inverse_loader,
  optimizer = inverse_optimizer,
  epochs = 25,
  n_components = 4,
  lambda = 0.05,
  verbose = FALSE,
  early_stop = TRUE,
  stop_window = 5,
  stop_delta = 1e-3
)
```

Given an observed polynomial output, we predict a posterior distribution over the original inputs.

```{r inverse-predict}
obs_test_scaled <- collapse::TRA(poly_test$output, attr(obs_scaled, "col_scale")$col_means, "-")
obs_test_scaled <- collapse::TRA(obs_test_scaled, attr(obs_scaled, "col_scale")$col_sds, "/")

inverse_pred <- probotPredictMDN(
  input = obs_test_scaled,
  model = inverse_fit$model,
  n_components = 4
)

inverse_summary <- probotMarginalPostMDN(
  mdn_output = inverse_pred,
  col_means = attr(param_scaled, "col_scale")$col_means,
  col_sds = attr(param_scaled, "col_scale")$col_sds,
  col_names = colnames(theta_train)
)

head(round(inverse_summary$post_mean, 3))
head(round(inverse_summary$post_sd, 3))
```

The posterior samples provide a direct way to inspect uncertainty and possible multimodality in the recovered inputs.

```{r inverse-samples}
post_samples <- probotSamplePostMDN(
  mdn_output = inverse_pred,
  galaxy_index = 1,
  n_samples = 2000,
  col_means = attr(param_scaled, "col_scale")$col_means,
  col_sds = attr(param_scaled, "col_scale")$col_sds,
  col_names = colnames(theta_train)
)

head(post_samples)
apply(post_samples, 2, sd)
```

Finally, we can compare the posterior mean with the true generating inputs for the held-out observations.

```{r inverse-metrics}
comparison <- cbind(
  truth = theta_test[1, ],
  estimate = inverse_summary$post_mean[1, ],
  sd = inverse_summary$post_sd[1, ]
)
comparison
```

This forward-and-inverse workflow captures the core ProBot interface:

- `probotScaleForward()` and `probotScaleBackward()` for standardisation.
- `probotDataLoader()` to assemble batched training data.
- `probotMakeMDN()` to define the MDN architecture.
- `probotTrainMDN()` to optimise the network.
- `probotPredictMDN()`, `probotMarginalPostMDN()`, and `probotSamplePostMDN()` to summarise and sample from the predictive posterior.
