---
title: "ProBot: Going with the Flow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ProBot: Going with the Flow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ProBot)
library(torch)
library(magicaxis)
set.seed(1)
torch_manual_seed(1)
eval_global <- FALSE
```

## Overview

**ProBot** supports two complementary approaches to simulation-based inference (SBI).
The *Back and Forth* vignette covers the classic **Mixture Density Network** (MDN)
workflow. This vignette introduces the second approach: **Normalising Flows** (NFs),
specifically a conditional Real-valued Non-Volume Preserving (RealNVP) architecture
built from affine coupling layers.

### What is a Normalising Flow?

A normalising flow learns an invertible, differentiable mapping between two probability
distributions. In the SBI context the flow maps a simple base distribution
(standard normal) to the complex posterior distribution over simulator parameters,
conditioned on observed data.

Concretely, the flow `f` transforms a latent vector `z ~ N(0, I)` into a parameter
vector `θ = f⁻¹(z | x)` for a given observation `x`. Training minimises the negative
log-likelihood of the true parameters under the induced distribution:

```
log p(θ | x) = log p_base(f(θ, x)) + log |det J_f(θ, x)|
```

The key advantage over MDNs is that the posterior is represented as an *exact*
density rather than a finite Gaussian mixture: the flow can represent arbitrarily
complex, multi-modal posteriors without fixing the number of components in advance.
Sampling is fast: draw `z ~ N(0, I)` and apply the inverse transformation.

### The coupling-layer architecture

**ProBot**'s `probotNetworkFlow` stacks multiple `probotCouplingLayer` modules.
Each coupling layer:

1. Splits the parameter vector `θ` into two halves, `θ₁` and `θ₂`.
2. Uses a small network that takes `(θ₁, x)` as input to predict an affine shift `t`
   and log-scale `s` for `θ₂`.
3. Applies the transformation `z₂ = (θ₂ − t) · exp(s)`, leaving `θ₁` unchanged.
4. Swaps the two halves so successive layers transform the complementary subset.

The log-determinant of the Jacobian is the sum of the log-scale values across
transformed dimensions — cheap to compute and numerically stable.

---

## The toy problem

We reuse the polynomial toy problem from the *Back and Forth* vignette. A generative
function maps two input parameters (`u`, `v`) to the coefficients of a polynomial
that is then evaluated on a nine-point grid, giving a nine-dimensional output vector.

```{r simulator}
make_polynomial_output <- function(theta) {
  a0 <- 0.3 + 0.8 * theta[, 1] - 0.4 * theta[, 2]
  a1 <- -1.2 + 0.5 * theta[, 1]^2 + 0.7 * theta[, 2]
  a2 <- 0.6 * sin(pi * theta[, 1]) - 0.3 * theta[, 2]^2

  x_grid <- seq(-1, 1, length.out = 9)
  y <- vapply(
    x_grid,
    function(x) a0 + a1 * x + a2 * x^2,
    numeric(nrow(theta))
  )

  colnames(y) <- paste0("x", seq_along(x_grid))
  list(output = y, x_grid = x_grid)
}
```

### Simulating training and test data

```{r simulate-data, eval = eval_global}
n_train <- 1e5
n_test  <- 1e4

theta_train <- cbind(
  u = runif(n_train, -1, 1),
  v = runif(n_train, -1, 1)
)

theta_test <- cbind(
  u = runif(n_test, -1, 1),
  v = runif(n_test, -1, 1)
)

poly_train <- make_polynomial_output(theta_train)
poly_test  <- make_polynomial_output(theta_test)
```

---

## Building a Normalising Flow for inverse inference

The inverse problem asks: given an observed nine-dimensional polynomial vector `x`,
what is the posterior distribution over the two generating parameters `(u, v)`?

### Scaling and data loading

As with MDNs, we standardise inputs and outputs before training.

```{r scale-data, eval = eval_global}
# The polynomial outputs are the context (conditioning variable)
output_scaled <- probotScaleForward(poly_train$output)

# The latent parameters are what the flow learns to recover
input_scaled  <- probotScaleForward(theta_train)
```

The flow is trained using `probotMultiEpochFlow()`, which leverages a `probotDataLoader`
to handle mini-batching and device placement efficiently.

```{r build-tensors, eval = eval_global}
flow_dataloader <- probotDataLoader(
  input  = output_scaled,
  output = input_scaled,
  batch  = 1024
)
```

### Constructing the flow network

`probotNetworkFlow` accepts three key hyperparameters:

- `dim_theta` — dimensionality of the parameter space (2 for `u` and `v`).
- `dim_x` — dimensionality of the conditioning context (9 polynomial values).
- `n_layers` — number of coupling layers; more layers increase expressiveness.
- `hidden_dim` — width of the shift/scale network inside each coupling layer.

```{r build-flow, eval = eval_global}
flow_model <- probotMakeFlow(
  dim_theta  = ncol(theta_train),
  dim_x      = ncol(poly_train$output),
  n_layers   = 6,
  hidden_dim = 64
)()
```

### Training the flow

Training is handled by `probotMultiEpochFlow()`, which wraps the negative-log-likelihood
loop and includes optional features like early stopping.

```{r train-flow, eval = eval_global}
flow_optimizer <- optim_adam(flow_model$parameters, lr = 3e-4)

train_result <- probotTrainFlow(
  model     = flow_model,
  dataloader = flow_dataloader,
  optimizer = flow_optimizer,
  epochs    = 100,
  verbose   = TRUE
)

flow_model <- train_result$model
loss_history <- train_result$history$loss
```

### Inspecting the training history

```{r plot-history, fig.width = 7, fig.height = 4, eval = eval_global}
magplot(
  seq_along(loss_history), loss_history,
  type = "l", col = "steelblue", lwd = 2,
  xlab = "Epoch", ylab = "Mean NLL",
  main = "Normalising Flow — training loss"
)
```

A monotonically decreasing loss that flattens toward the end confirms the flow has
converged. Unlike an MDN, the NLL of a flow does not have the same units across
problems, but steady decrease and eventual plateau are the key signals.

---

## Posterior sampling

Once trained, drawing samples from the posterior for a new observation requires only:

1. Draw `z ~ N(0, I)` (one sample per desired posterior point).
2. Evaluate the inverse flow: `θ = f⁻¹(z | x_obs)`.
3. Unscale the result back to the original parameter space.

`probotSamplePostNF()` encapsulates these steps.

```{r scale-test, eval = eval_global}
# Scale the test context with the training statistics
output_test_scaled <- probotScaleForward(
  poly_test$output,
  col_means = attr(output_scaled, "col_scale")$col_means,
  col_sds   = attr(output_scaled, "col_scale")$col_sds
)
```

### Posterior for a single observation

```{r single-posterior, eval = eval_global}
obs_idx <- 1
context_obs <- torch_tensor(
  output_test_scaled[obs_idx, , drop = FALSE],
  dtype = torch_float()
)$to(device = torch_device('mps'))

post_samples <- probotSamplePostNF(
  context_x  = context_obs,
  model      = flow_model,
  n_samples  = 5000,
  col_means  = attr(input_scaled, "col_scale")$col_means,
  col_sds    = attr(input_scaled, "col_scale")$col_sds,
  col_names  = colnames(theta_train)
)

head(post_samples)
```

The corner plot below shows the joint posterior over `(u, v)` for the first test
observation. The blue reference lines mark the true generating values.

```{r corner-plot, fig.width = 5, fig.height = 5, eval = eval_global}
magtri(post_samples, refvals = theta_test[obs_idx, ],
       quan_range = c(0.02, 0.98))
```

Because the forward mapping from `(u, v)` to the nine polynomial values is nonlinear
and non-injective, the posterior can be multimodal: two distinct parameter settings
can produce nearly identical polynomial curves. The normalising flow captures this
multi-modal structure naturally without specifying the number of modes in advance.

---

## Population-level summaries

For a broader assessment we compute posterior means and standard deviations across
the test set by sampling the flow at each observation.

```{r population-summaries, eval = eval_global}
n_eval    <- min(n_test, 2000)   # limit for speed in this demo
post_mean <- matrix(NA_real_, n_eval, ncol(theta_test), dimnames = list(NULL, colnames(theta_test)))
post_sd   <- matrix(NA_real_, n_eval, ncol(theta_test), dimnames = list(NULL, colnames(theta_test)))

for (i in seq_len(n_eval)) {
  ctx_i <- torch_tensor(
    output_test_scaled[i, , drop = FALSE],
    dtype = torch_float()
  )$to(device = torch_device('mps'))

  samp_i <- probotSamplePostNF(
    context_x = ctx_i,
    model     = flow_model,
    n_samples = 500,
    col_means = attr(input_scaled, "col_scale")$col_means,
    col_sds   = attr(input_scaled, "col_scale")$col_sds,
    col_names = colnames(theta_train)
  )

  post_mean[i, ] <- colMeans(samp_i)
  post_sd[i, ]   <- apply(samp_i, 2, sd)

  if (i %% 500 == 0) cat(i, "\n")
}
```

### Truth vs. predicted scatter plots

For each parameter we plot the posterior mean against the true generating value.
Tight, symmetric scatter around the diagonal confirms accurate posterior location.

```{r scatter-plots, eval = eval_global}
for (j in seq_len(ncol(theta_test))) {
  magplot(
    theta_test[seq_len(n_eval), j], post_mean[, j],
    xlab = paste0("True ",      colnames(theta_test)[j]),
    ylab = paste0("Predicted ", colnames(theta_test)[j]),
    pch  = "."
  )
  magerr(
    theta_test[seq_len(n_eval), j], post_mean[, j],
    ylo = post_sd[, j],
    col = hsv(v = 0, alpha = 0.05)
  )
  abline(0, 1, col = "red", lty = 2)
}
```

---

## Calibration diagnostics

Because `probotPIT()` and `probotTARP()` call `probotSamplePostMDN()` internally,
we implement equivalent NF-based diagnostics directly. The logic is identical — only
the sampling function differs.

### PIT: marginal calibration

For each test observation and each parameter dimension the PIT value is the fraction
of posterior samples that fall *below* the true value. Under a perfect posterior
this is uniform on [0, 1].

```{r pit-nf, eval = eval_global}
n_pit   <- min(n_test, 1000)
n_samp_pit <- 2000

pit_nf <- matrix(NA_real_, n_pit, ncol(theta_test),
                 dimnames = list(NULL, colnames(theta_test)))

for (i in seq_len(n_pit)) {
  ctx_i <- torch_tensor(
    output_test_scaled[i, , drop = FALSE],
    dtype = torch_float()
  )$to(device = torch_device('mps'))

  samp_i <- probotSamplePostNF(
    context_x = ctx_i,
    model     = flow_model,
    n_samples = n_samp_pit,
    col_means = attr(input_scaled, "col_scale")$col_means,
    col_sds   = attr(input_scaled, "col_scale")$col_sds,
    col_names = colnames(theta_train)
  )

  for (j in seq_len(ncol(theta_test))) {
    pit_nf[i, j] <- mean(samp_i[, j] <= theta_test[i, j])
  }

  if (i %% 200 == 0) cat(i, "\n")
}
```

```{r pit-plot, fig.width = 5, fig.height = 5, eval = eval_global}
magtri(pit_nf)
```

**What to look for:**

- *Uniform histograms* on the diagonal → good marginal calibration.
- *U-shaped histograms* → posterior too narrow (truth falls in the tails too often).
- *Λ-shaped histograms* → posterior too wide (truth clusters near the centre).
- *Skewed histograms* → systematic bias in the posterior location.

### TARP: joint calibration

TARP projects posterior samples and the true parameter vector onto a random unit
direction and computes the coverage fraction. Repeating over many observations and
averaging over random directions gives an omnibus test of the full joint posterior.

```{r tarp-nf, eval = eval_global}
n_tarp      <- min(n_test, 1000)
n_samp_tarp <- 2000

tarp_nf <- numeric(n_tarp)

for (i in seq_len(n_tarp)) {
  ctx_i <- torch_tensor(
    output_test_scaled[i, , drop = FALSE],
    dtype = torch_float()
  )$to(device = device)

  samp_i <- probotSamplePostNF(
    context_x = ctx_i,
    model     = flow_model,
    n_samples = n_samp_tarp,
    col_means = attr(input_scaled, "col_scale")$col_means,
    col_sds   = attr(input_scaled, "col_scale")$col_sds,
    col_names = colnames(theta_train)
  )

  direction    <- rnorm(ncol(samp_i))
  direction    <- direction / sqrt(sum(direction^2))
  truth_proj   <- sum(theta_test[i, ] * direction)
  sample_proj  <- as.vector(samp_i %*% direction)
  tarp_nf[i]  <- mean(sample_proj <= truth_proj)

  if (i %% 200 == 0) cat(i, "\n")
}
```

```{r tarp-plot, fig.width = 6, fig.height = 4, eval = eval_global}
maghist(
  tarp_nf, breaks = 20,
  xlab = "TARP coverage", ylab = "Count",
  main = "Normalising Flow — TARP diagnostic"
)
abline(h = length(tarp_nf) / 20, lty = 2, col = "red")

ks.test(tarp_nf, "punif")
```

**Interpreting the TARP histogram:**

- *Flat histogram* → joint posterior is well calibrated.
- *Excess mass near 0.5* → posterior is over-dispersed (too wide in joint space).
- *Excess mass near 0 and 1* → posterior is under-dispersed (too narrow in joint space).
- *Significant KS p-value* (p < 0.05) → evidence of joint miscalibration.

---

## Normalising Flow vs. MDN: when to use which

| Criterion | MDN | Normalising Flow |
|---|---|---|
| Posterior representation | Finite Gaussian mixture | Exact, arbitrarily complex |
| Multi-modality | Limited to K components | Unlimited (architecture permitting) |
| Training stability | Generally stable | May require careful tuning |
| Sampling speed | Very fast (closed-form mixture) | Fast (single inverse pass) |
| Density evaluation | Tractable mixture sum | Tractable via change-of-variables |
| Typical use case | Low-to-medium posterior complexity | Complex / multi-modal posteriors |

Normalising flows are particularly well-suited when:

- The true posterior is known to be multi-modal or banana-shaped.
- A fixed number of Gaussian components cannot capture the posterior adequately.
- You need a continuous, differentiable posterior density (e.g., for nested sampling seeding).

MDNs are often preferable when:

- The posterior is approximately Gaussian or a small mixture thereof.
- Training speed and simplicity are priorities.
- You want built-in ProBot utilities (e.g., `probotPIT()`, `probotTARP()`) without modification.

---

## Summary

| Function | Role |
|---|---|
| `probotCouplingLayer` | Single affine coupling layer: maps `(θ, x) → (z, log_det_jac)` |
| `probotNetworkFlow` | Stacked coupling-layer flow: forward (θ → z) and inverse (z → θ) |
| `probotLossNF()` | Negative log-likelihood loss for flow training |
| `probotSamplePostNF()` | Draw posterior samples via the inverse flow |
| `probotScaleForward()` / `probotScaleBackward()` | Standardise columns and reverse the transform |

The normalised flow architecture adds an expressive alternative to the MDN backbone
without changing the broader ProBot workflow: scale data, train on (parameter,
observation) pairs, sample the posterior, validate calibration with PIT and TARP.
The key difference is that the posterior is now represented as an invertible
transformation of a standard normal, rather than a finite mixture of Gaussians —
enabling exact density evaluation and richer posteriors at the cost of a slightly
more involved training loop.
